On the uniqueness and stability of dictionaries \\ for sparse representation of noisy signals

### 2019-05-09 manuscripts/IEEE/final_submission.tex
tidying up

%

\begin{problem}\label{InverseProblem}
Find a real matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$ that satisfy $\|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta$ for all $i$.
\end{problem}

@@

We generalize the matrix lower bound for our purposes by restricting it to the spans of certain submatrices\footnote{See \cite{vidal2005generalized} for an overview of the related ``union of subspaces" model.} associated with a hypergraph $\mathcal{H} \subseteq {[m] \choose k}$ of column indices. Let $\mathbf{M}_S$ denote the submatrix formed by the columns of a matrix $\mathbf{M}$ indexed by $S \subseteq [m]$ (setting $\mathbf{M}_\emptyset := \mathbf{0}$).  In the sections that follow, we shall also let $\bm{\mathcal{M}}_S$ denote the column-span of a submatrix $\mathbf{M}_S$, and $\bm{\mathcal{M}}_\mathcal{G}$ to denote $\{\bm{\mathcal{M}}_S\}_{S \in \mathcal{G}}$.  We define:  % Define $L_\mathcal{H}(\mathbf{M})$ as follows:
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \ falpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \mathbf{0} \neq \mathbf{x} \in \mathbb{R}^{|S|} \right\},
\end{align} 
%
writing also $L_{k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{In compressed sensing literature, \mbox{$1 - \sqrt{k} L_k(\mathbf{M})$}  is the asymmetric lower restricted isometry constant for $\mathbf{M}$ with unit $\ell_2$-norm columns \cite{Blanchard2011}.}  As explained above, compactness implies that $L_{2k}(\mathbf{M}) > 0$ for all $\mathbf{M}$ satisfying \eqref{SparkCondition}. Clearly, $L_{\mathcal{H}'}(\mathbf{M}) \geq L_\mathcal{H}(\mathbf{M})$ whenever $\mathcal{H}' \subseteq \mathcal{H}$, and similarly any $k$-uniform $\mathcal{H}$ satisfying $\cup \mathcal{H} = [m]$ has $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$ (letting $L_{2k}$ := $L_m$ whenever $2k > m$).

@@

Since sparse codes in general linear position are straightforward to produce deterministically with a ``Vandermonde''  construction (i.e., by choosing columns of the matrix $[\gamma_{i}^j]_{i,j=1}^{k,N}$, for distinct nonzero $\gamma_i$), we have the following direct consequence of Thm.~\ref{DeterministicUniquenessTheorem}.

@@

%\begin{align*}
%\|\mathbf{BP}\mathbf{x}\|_2 
%&\geq \|\mathbf{A}\mathbf{D}^{-1}\mathbf{x}\|_2 - \|(\mathbf{A} - \mathbf{BPD})\mathbf{D}^{-1}\mathbf{x}\|_2 \\
%&\geq \sqrt{2k} (L_{2k}(\mathbf{A}) -  C_1\varepsilon) \|\mathbf{D}^{-1}\mathbf{x}\|_2, \\ 
%&\geq \sqrt{2k} (L_{2k}(\mathbf{A}) -  C_1\varepsilon) \|\mathbf{D}^{-1}\|_{?} \| \mathbf{x}\|_2,
%\end{align*}
%
We therefore have that $L_{2k}(\mathbf{BPD}) \geq L_{2k}(\mathbf{A}) - C_1\varepsilon  > 0$, and \eqref{b-PDa} then follows from \eqref{stuff}. The reader may verify that $L_{2k}(\mathbf{BP}) \geq L_{2k}(\mathbf{BPD}) / \|\mathbf{D}\|_1$.

@@

We now state our result in combinatorial matrix analysis, which contains most of the complexity in the proof of Thm.~\ref{DeterministicUniquenessTheorem}.

@@

The constant $C_2 > 0$ of Lem.~\ref{MainLemma} is then given by:  
\begin{align}\label{Cdef2}
	C_2(\mathbf{A}, \mathcal{H}) := \frac{ (r+1) \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ 1- \max_{\mathcal{G} \in {\mathcal{H} \choose r+1}} \xi( \bm{\mathcal{A}}_\mathcal{G} ) }.
\end{align}

We now define the constant $C_1 > 0$ of Thm.~\ref{DeterministicUniquenessTheorem} in terms of $C_2$. Given vectors $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$, let $\mathbf{X}$ denote the $m \times N$ matrix with columns $\mathbf{x}_i$ and let $I(S)$ denote the set of indices $i$ for which $\mathbf{x}_i$ is supported in $S$.  We define:
\begin{align}\label{Cdef1}
C_1(\mathbf{A}, \mathcal{H}, \{\mathbf{x}_i\}_{i=1}^N) := \frac{ C_2(\mathbf{A}, \mathcal{H}) } { \min_{S \in \mathcal{H}} L_k(\mathbf{AX}_{I(S)}) }.
\end{align}
Given the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} on $\mathbf{A}$ and the $\mathbf{x}_i$, this expression for $C_1$ is well-defined\footnote{\label{note1}To see this, fix $S \in \mathcal{H}$ and $k$-sparse $\mathbf{c}$. Using the definitions, we have $\|\mathbf{AX}_{I(S)}\mathbf{c}\|_2 \geq \sqrt{k} L_\mathcal{H}(\mathbf{A})\|\mathbf{X}_{I(S)}\mathbf{c}\|_2 \geq k L_\mathcal{H}(\mathbf{A}) L_k(\mathbf{X}_{I(S)})\|\mathbf{c}\|_2$. Thus, $L_k(\mathbf{AX}_{I(S)}) \geq \sqrt{k} L_\mathcal{H}(\mathbf{A}) L_k(\mathbf{X}_{I(S)}) > 0$, since $L_{\mathcal{H}}(\mathbf{A}) \geq L_{2\mathcal{H}}(\mathbf{A})> 0$ and $L_k(\mathbf{X}_{I(S)}) > 0$ by general linear position of the $\mathbf{x}_i$.} and yields an upper bound on $\varepsilon$ consistent with that proven sufficient in the case $k=1$ considered at the beginning of this section.\footnote{When $\mathbf{x}_i = c_i\mathbf{e}_i$ we have $C_2 \geq 2\|\mathbf{A}_i\|_2$ and the denominator in \eqref{Cdef1} becomes $\min_{i \in [m]} |c_i| \|\mathbf{A}_i\|_2$, hence $C_1 \geq 2 / \min_{i \in [m]} |c_i|$.}
% C_2(\mathbf{A}, \mathcal{H}) := \frac{ (r+1) \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ \min_{\mathcal{G} \in {\mathcal{H} \choose r} \cup {\mathcal{H} \choose r+1}} \xi( \{ \bm{\mathcal{A}}_S \}

>>>>

\begin{problem}\label{InverseProblem}
Find a real matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$ that satisfy $\|\mathbf{z}_i - \mathbf{B}\mathbf{\overline x}_i\|_2 \leq \eta$ for all $i = 1,\ldots,N$.
\end{problem}

@@

The lower bound of a matrix is generalized below in (\ref{Ldef}) by restricting it to the spans of certain submatrices\footnote{See \cite{vidal2005generalized} for an overview of the related ``union of subspaces" model.} associated with a hypergraph $\mathcal{H} \subseteq {[m] \choose k}$ of column indices. Let $\mathbf{M}_S$ denote the submatrix formed by the columns of a matrix $\mathbf{M}$ indexed by $S \subseteq [m]$ (setting $\mathbf{M}_\emptyset := \mathbf{0}$).  In the sections that follow, we shall also let $\bm{\mathcal{M}}_S$ denote the column-span of a submatrix $\mathbf{M}_S$, and $\bm{\mathcal{M}}_\mathcal{G}$ to denote $\{\bm{\mathcal{M}}_S\}_{S \in \mathcal{G}}$.  We define:  % Define $L_\mathcal{H}(\mathbf{M})$ as follows:
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \ falpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \mathbf{0} \neq \mathbf{x} \in \mathbb{R}^{|S|} \right\},
\end{align} 
%
writing also $L_{k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{In compressed sensing literature, \mbox{$1 - \sqrt{k} L_k(\mathbf{M})$}  is the asymmetric lower restricted isometry constant for $\mathbf{M}$ with unit $\ell_2$-norm columns \cite{Blanchard2011}.}  As explained above, compactness implies that $L_{2k}(\mathbf{M}) > 0$ for all $\mathbf{M}$ satisfying \eqref{SparkCondition}. Clearly, $L_{\mathcal{H}'}(\mathbf{M}) \geq L_\mathcal{H}(\mathbf{M})$ whenever $\mathcal{H}' \subseteq \mathcal{H}$, and similarly any $k$-uniform $\mathcal{H}$ satisfying $\cup \mathcal{H} = [m]$ has $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$ (letting $L_{2k}$ := $L_m$ whenever $2k > m$).

@@

Since sparse codes in general linear position are straightforward to produce with a ``Vandermonde''  construction (i.e., by choosing columns of the matrix $[\gamma_{i}^j]_{i,j=1}^{k,N}$, for distinct nonzero $\gamma_i$), we have the following direct consequence of Thm.~\ref{DeterministicUniquenessTheorem}.

@@

%\begin{align*}
%\|\mathbf{BP}\mathbf{x}\|_2 
%&\geq \|\mathbf{A}\mathbf{D}^{-1}\mathbf{x}\|_2 - \|(\mathbf{A} - \mathbf{BPD})\mathbf{D}^{-1}\mathbf{x}\|_2 \\
%&\geq \sqrt{2k} (L_{2k}(\mathbf{A}) -  C_1\varepsilon) \|\mathbf{D}^{-1}\mathbf{x}\|_2, \\ 
%&\geq \sqrt{2k} (L_{2k}(\mathbf{A}) -  C_1\varepsilon) \|\mathbf{D}^{-1}\|_{?} \| \mathbf{x}\|_2,
%\end{align*}
%
We therefore have that $L_{2k}(\mathbf{BPD}) \geq L_{2k}(\mathbf{A}) - C_1\varepsilon  > 0$, and \eqref{b-PDa} then follows from \eqref{stuff}. The reader may also verify that $L_{2k}(\mathbf{BP}) \geq L_{2k}(\mathbf{BPD}) / \|\mathbf{D}\|_1$.

@@

The following is our result in combinatorial matrix analysis; it contains most of the complexity in the proof of Thm.~\ref{DeterministicUniquenessTheorem}.

@@

The constant $C_2 > 0$ of Lem.~\ref{MainLemma} can now be expressed as:  
\begin{align}\label{Cdef2}
	C_2(\mathbf{A}, \mathcal{H}) := \frac{ (r+1) \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ 1- \max_{\mathcal{G} \in {\mathcal{H} \choose r+1}} \xi( \bm{\mathcal{A}}_\mathcal{G} ) }.
\end{align}

We next define the constant $C_1 > 0$ of Thm.~\ref{DeterministicUniquenessTheorem} in terms of $C_2$. Given vectors $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$, let $\mathbf{X}$ denote the $m \times N$ matrix with columns $\mathbf{x}_i$ and let $I(S)$ denote the set of indices $i$ for which $\mathbf{x}_i$ is supported in $S$.  We define:
\begin{align}\label{Cdef1}
C_1(\mathbf{A}, \mathcal{H}, \{\mathbf{x}_i\}_{i=1}^N) := \frac{ C_2(\mathbf{A}, \mathcal{H}) } { \min_{S \in \mathcal{H}} L_k(\mathbf{AX}_{I(S)}) }.
\end{align}
Given the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} on $\mathbf{A}$ and the $\mathbf{x}_i$, this expression for $C_1$ is well-defined\footnote{\label{note1}To see this, fix $S \in \mathcal{H}$ and $k$-sparse $\mathbf{c}$. Using the definitions, we have $\|\mathbf{AX}_{I(S)}\mathbf{c}\|_2 \geq \sqrt{k} L_\mathcal{H}(\mathbf{A})\|\mathbf{X}_{I(S)}\mathbf{c}\|_2 \geq k L_\mathcal{H}(\mathbf{A}) L_k(\mathbf{X}_{I(S)})\|\mathbf{c}\|_2$. Thus, $L_k(\mathbf{AX}_{I(S)}) \geq \sqrt{k} L_\mathcal{H}(\mathbf{A}) L_k(\mathbf{X}_{I(S)}) > 0$, since $L_{\mathcal{H}}(\mathbf{A}) \geq L_{2\mathcal{H}}(\mathbf{A})> 0$ and $L_k(\mathbf{X}_{I(S)}) > 0$ by general linear position of the $\mathbf{x}_i$.} and yields an upper bound on $\varepsilon$ consistent with that proven sufficient in the case $k=1$ considered at the beginning of this section.\footnote{When $\mathbf{x}_i = c_i\mathbf{e}_i$, we have $C_2 \geq 2\|\mathbf{A}_i\|_2$ and the denominator in \eqref{Cdef1} becomes $\min_{i \in [m]} |c_i| \|\mathbf{A}_i\|_2$; hence, $C_1 \geq 2 / \min_{i \in [m]} |c_i|$.}
% C_2(\mathbf{A}, \mathcal{H}) := \frac{ (r+1) \max_{j \in [m]} \|\mathbf{A}_j\|_2}{ \min_{\mathcal{G} \in {\mathcal{H} \choose r} \cup {\mathcal{H} \choose r+1}} \xi( \{ \bm{\mathcal{A}}_S \}
